---
title: "Biostat 212a Homework 5"
subtitle: "Due Mar 11, 2024 @ 11:59PM"
author: "Chengwu Duan (Jason) and 606332825"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

## ISL Exercise 9.7.1 (10pts)

This problem involves hyperplanes in two dimensions.

(a) Sketch the hyperplane $1 + 3X_1 − X_2 = 0$. Indicate the set of
points for which $1 + 3X_1 − X_2 > 0$, as well as the set of points
for which $1 + 3X_1 − X_2 < 0$.

```{r}
library(ggplot2)

# Define the range for X1 and X2
x1_range <- seq(-10, 10, length.out = 300)
x2_range <- seq(-30, 30, length.out = 300)

# Create a grid of X1, X2 values
grid <- expand.grid(X1 = x1_range, X2 = x2_range)

# Calculate 1 + 3X1 - X2
grid$hyperplane_value <- with(grid, 1 + 3 * X1 - X2)

# Base plot with points colored based on the hyperplane value
ggplot(grid, aes(x = X1, y = X2, fill = hyperplane_value)) +
  geom_tile() + # Use geom_tile to color areas based on hyperplane_value
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-10, 10), space = "Lab", 
                       name="Hyperplane\nValue") +
  geom_contour(aes(z = hyperplane_value), binwidth = 1, colour = "black") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(title = "Hyperplane 1 + 3X1 - X2 = 0",
       x = expression(X[1]),
       y = expression(X[2])) +
  theme_minimal() 
```

**Answer: **
The area above the line, colored in shades of blue, represents the set of points for which $1 + 3X_1 - X_2 < 0$ . This inequality holds true for points on the plane that are below the hyperplane line because the actual value of $X_2$ for these points is greater than $1 + 3X_1$. 

The area below the line, colored in shades of red, represents the set of points for which $1 + 3X_1 - X_2 > 0$. This inequality holds true for points on the plane that are above the hyperplane line because the actual value of $X_2$ for these points is less than $1 + 3X_1$.

The color intensity reflects the magnitude of the value $1 + 3X_1 + X_2$: darker shades indicate values farther from zero (either positive or negative), while lighter shades are closer to zero. The hyperplane itself is represented by the region where the color transitions from blue to red, indicating the value is exactly zero at the line. Points are classified based on the region where it lies in being above the line are classified as positive, while points below the line are classified as negative.

(b) On the same plot, sketch the hyperplane $−2 + X_1 + 2X_2 = 0$.
Indicate the set of points for which $−2 + X_1 + 2X_2 > 0$, as well
as the set of points for which $−2 + X_1 + 2X_2 < 0$.

```{r}
library(ggplot2)

# Define the range for X1 and X2
x1_range <- seq(-10, 10, length.out = 300)
x2_range <- seq(-30, 30, length.out = 300)

# Create a grid of X1, X2 values
grid <- expand.grid(X1 = x1_range, X2 = x2_range)

# Calculate 1 + 3X1 - X2
grid$hyperplane_value <- with(grid, -2 + X1 + 2 * X2)

# Base plot with points colored based on the hyperplane value
ggplot(grid, aes(x = X1, y = X2, fill = hyperplane_value)) +
  geom_tile() + # Use geom_tile to color areas based on hyperplane_value
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-10, 10), space = "Lab", 
                       name="Hyperplane\nValue") +
  geom_contour(aes(z = hyperplane_value), binwidth = 1, colour = "black") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(title = "Hyperplane -2 + X1 + 2X2 = 0",
       x = expression(X[1]),
       y = expression(X[2])) +
  theme_minimal() 
```

**Answer: ** same logic as part a.

## ISL Exercise 9.7.2 (10pts)

We have seen that in p = 2 dimensions, a linear decision boundary
takes the form $β_0 +β_1X_1 +β_2X_2 = 0$. We now investigate a non-linear
decision boundary.

(a) Sketch the curve
$(1 + X_1)^2 + (2 − X_2)^2 = 4$.

```{r}
library(ggplot2)

# Function to calculate the circle's radius squared
radius_squared <- function(x, y) {
  (1 + x)^2 + (2 - y)^2
}

# Define the range for X1 and X2 and create a data frame
x_range <- seq(-4, 2, length.out = 300)
y_range <- seq(0, 4, length.out = 300)
df <- expand.grid(X1 = x_range, X2 = y_range)

# Calculate the Z values (the radius squared for each point)
df$Z <- mapply(radius_squared, df$X1, df$X2)

# Plot the contour
ggplot(df, aes(X1, X2, z = Z)) +
  stat_contour(breaks = 4, colour = "blue") +
  scale_x_continuous(name = "X1") +
  scale_y_continuous(name = "X2") +
  ggtitle("Contour of (1 + X1)^2 + (2 - X2)^2 = 4") +
  theme_minimal()
```

(b) On your sketch, indicate the set of points for which
$(1 + X_1)^2 + (2 − X_2)^2 > 4$,

as well as the set of points for which $(1 + X_1)^2 + (2 − X_2)^2 ≤ 4$.

(c) Suppose that a classifier assigns an observation to the blue class
if
$(1 + X_1)^2 + (2 − X_2)^2 > 4$,
and to the red class otherwise. To what class is the observation
(0, 0) classified? (−1, 1)? (2, 2)? (3, 8)?

(d) Argue that while the decision boundary in (c) is not linear in
terms of $X_1 and X_2$, it is linear in terms of $X_1, X_1^2 , X_2, and X_2^2$. 

## Support vector machines (SVMs) on the `Carseats` data set (30pts)

Follow the machine learning workflow to train support vector classifier (same as SVM with linear kernel), SVM with polynomial kernel (tune the degree and regularization parameter $C$), and SVM with radial kernel (tune the scale parameter $\gamma$ and regularization parameter $C$) for classifying `Sales<=8` versus `Sales>8`. Use the same seed as in your HW4 for the initial test/train split and compare the final test AUC and accuracy to those methods you tried in HW4.

## Bonus (10pts)

Let
$$
f(X) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p = \beta_0 + \beta^T X. 
$$
Then $f(X)=0$ defines a hyperplane in $\mathbb{R}^p$. Show that $f(x)$ is proportional to the signed distance of a point $x$ to the hyperplane $f(X) = 0$. 
